### 2019-02-21 chapter_00_open_science.tex
Add todo notes and cleanup chapter 00

\chapter{Towards reproducible, open science through open research software}

\section*{(1) Overview}

\vspace{0.5cm}

\section*{Title}

@@

Taking advantage of recent advances in open-source software engineering to study
fluid dynamics: the example of the FluidDyn project

\section*{Abstract}

% \textcolor{blue}{A short (ca. 100 word) summary of the software being
% described: what problem the software addresses, how it was implemented and
% architected, where it is stored, and its reuse potential.}

\href{http://fluiddyn.readthedocs.io}{FluidDyn} is a project to foster
open-science and open-source in the fluid dynamics community.  It is thought of
as a research project to channel open-source dynamics, methods and tools to do
better science.
%
We propose a set of Python packages forming a framework to study fluid dynamics
with different methods, in particular laboratory experiments (package
\fluidpack{lab}), simulations (packages \fluidpack{fft}, \fluidpack{sim} and
\fluidpack{foam}) and data processing (package \fluidpack{image}).
%
In the present article, we review the recent advancements in web-based software
development tools and Python which facilitated the execution of this project,
and establish a picture of software engineering in fluid mechanics.
%
We give an overview of the specialized packages of the project and then
focus on the base package called \fluidpack{dyn}, which contains common code
used in the specialized packages.  Packages \fluidpack{fft} and \fluidpack{sim}
are described with greater detail in two companion papers, \citet{fluidfft,
fluidsim}.
%
With the project FluidDyn, we demonstrate that specialized scientific code can
be written with methods and good practices of the open-source community. The
Mercurial repositories are available in Bitbucket
(\url{https://bitbucket.org/fluiddyn/}). All codes are documented using Sphinx
and Read the Docs, and tested with continuous integration run on Bitbucket
Pipelines and Travis.
%
To improve the reuse potential, the codes are as modular as possible, leveraging
the simple object-oriented programming model of Python.
%
All codes are also written to be highly efficient, using C++, Cython and
Pythran to speedup the performance of critical functions.

\section*{Keywords}

% \textcolor{blue}{keyword 1; keyword 2; etc. \\
% Keywords should make it easy to identify who and what the software will be
% useful for.}

Fluid dynamics research with Python. Free and open-source software, modular,
object-oriented, collaborative, efficient, tested, documented.

\section*{Introduction}

@@

\subsection*{Science, software, open-source and the computer revolution}

@@

The section ``Overview'' is completed by a second subsection called
``Implementation and architecture'' devoted to present the organization of the
packages of the project and to describe in more detail the base package
\fluiddyn\footnote{We use FluidDyn (with capital letters) to name the project and
\fluiddyn for the base package.}.

\subsection*{Methods and tools for open-source software engineering}

@@

\subsection*{Python, a programming language adapted for open-science}

@@

\subsubsection*{An incredible success}

@@

\subsubsection*{Some Python issues}

@@

\subsubsection*{Rich landscape of open-source languages useful for science}

@@

\subsubsection*{Summary: Python, a versatile Swiss Army Knife for the
scientists}

@@

\subsection*{Productivity at individual, group and community levels}

@@

\subsection*{Programming in the field of fluid mechanics}

@@

\subsubsection*{Discussion on Matlab, Labview and Mathematica}

@@

\subsubsection*{Different models for software development in fluid mechanics}

@@

\subsection*{FluidDyn project: Breaking the status quo}

@@

\section*{Implementation and architecture}

@@

\subsection*{Organization of the code in packages}

@@

\subsection*{API of the Python library \fluiddyn}

@@

Sub-packages \codeinline{io}, \codeinline{util} and \codeinline{calcul} are the
largest in terms of lines of code and provides the Application Programming
Interfaces (API) to support \fluidpack{sim}, \fluidpack{lab} and
\fluidpack{image}.
%
For the sake of brevity, we shall only describe here some of the most important
modules.

\subsubsection*{Module \codeinline{fluiddyn.util.paramcontainer}}

Sub-package \codeinline{paramcontainer} defines class
\codeinline{ParamContainer} which is a hierarchical container for any type of
parameters. As shown in
\href{http://fluiddyn.readthedocs.io/en/latest/ipynb/tuto_paramscontainer.html}{%
this tutorial}, various strengths of an object of this class include:
\begin{itemize}
\item Support containing and printing documentation on the parameters.
\item Support printing to console as XML and saving as XML, hdf5 and netcdf4
files.
\item Evaluate data types automatically while loading from saved files.
\item Easy exploration in an interactive console.
\item Allows modification of default parameters through simple Python script
files. An error is raised if the user attempts to use an invalid parameter.
\item Graphical User Interface (GUI) frontend with PyQt.
\end{itemize}

Thus, it makes it a much more robust implementation for saving key parameters,
compared to conventional methods which rely on text or CSV files.

\subsubsection*{Module \codeinline{fluiddyn.util.serieofarrays}}

This modules provides classes to iterate over files. It is a common task in data
processing to understand how to organized a \textit{serie} of files from filenames
and formats. Then one has to form smaller sets of arrays contained in the
files. For example, we can have such \textit{serie}: \codeinline{im1\_1.png,
im1\_2.png, im1\_3.png, im2\_1.png, im2\_2.png, im2\_3.png} and we could create
subsets like \codeinline{((im1\_1.png, im1\_2.png, im1\_2.png), (im2\_1.png,
im2\_2.png, im2\_3.png))}. From the same \textit{serie} of images, we can also
create other subsets \codeinline{((im1\_1.png, im2\_1.png), (im1\_2.png,
im2\_2.png), (im1\_3.png, im2\_3.png))}.

The classes of this module allows one to do it with a quite simple and general
API as shown in
\href{http://fluiddyn.readthedocs.io/en/latest/ipynb/tuto_serieofarrays.html}{%
the tutorial on the module}.

\subsubsection*{Module \codeinline{fluiddyn.util.mpi}}

This simple module makes simulataneous sequential and MPI programming a breeze
by providing number of processes, \codeinline{nb\_proc = 1} and \codeinline{rank
= 0}, when used in sequential mode, otherwise providing the appropriate values
provided by \pack{mpi4py} package. Use of this module thwarts coding several if-else
clauses. If the program was using MPI, also defines the variable
\codeinline{comm} as an alias for the \codeinline{MPI.COMM\_WORLD} communicator.

\subsubsection*{Module \codeinline{fluiddyn.calcul.easyfft}}

Thin wrapper for an unified API using classes around the packages \pack{pyfftw}
and \pack{scipy.fftpack}.  It is very easy to perform forward and inverse Fast
Fourier Transforms (FFT) in one-, two- and three-dimensions.
The FFT can be multithreaded if the environment variable
\codeinline{OMP\_NUM\_THREADS} is defined.

\subsection*{The \fluiddyn command-line utilities}

The package \fluiddyn also provides few command-line utilities to perform
simple tasks useful for a scientist developing with Python and using the
FluidDyn packages.
\begin{itemize}

\item \codeinline{fluidinfo}

Displays important information related to software and hardware. It
includes detailed information such as currently installed FluidDyn packages,
other third-party packages, C compiler, MPI and \Numpy configuration.

\item \codeinline{fluiddump}

Utility to print the hierarchy of HDF5 and NetCDF files. It does not depend on
the NetCDF4 library.

\item \codeinline{fluidnbstripout}

Very simple layer to stripout Jupyter notebooks of output, which is useful to
keep the notebooks lightweight when included in a repository.
%
This tool is based on \href{https://github.com/kynan/nbstripout}{nbstripout}
but, in contrast to nbstripout,
%
by default the notebooks with a file name ending as `.nbconvert.ipynb' are
excluded.

\item \codeinline{fluidmat2py}

Utility to produce a strange code which is no longer Matlab and not yet Python.
%
This strange code is then much easier to translate into correct Python than the
original Matlab code.

\section*{Quality control}

% \textcolor{blue}{Detail the level of testing that has been carried out on the
% code (e.g. unit, functional, load etc.), and in which environments. If not
% already included in the software documentation, provide details of how a user
% could quickly understand if the software is working (e.g. providing examples
% of running the software with sample input and output data). }

The package \fluiddyn\ currently supplies unit tests covering around 70\% of its
code.  These unit tests are run regularly through continuous integration on Travis
CI with the most recent releases of \fluiddyn's dependencies and on Bitbucket
Pipelines inside a static \href{https://hub.docker.com/u/fluiddyn}{Docker
container}.  The tests are run using standard Python interpreter with all
supported versions.

For \fluiddyn, the code coverage results are displayed at
\href{https://codecov.io/gh/fluiddyn/fluiddyn}{Codecov}.  Using third-party
packages \pack{coverage} and \pack{tox}, it is straightforward to bootstrap the
installation with dependencies, test with multiple Python versions and combine the
code coverage report, ready for upload. It is also possible to run similar
isolated tests using \pack{tox} or coverage analysis using \pack{coverage} in a
local machine.  Up-to-date build status and coverage status are displayed on the
landing page of the Bitbucket repository.

We also try to follow a consistent code style as recomended by PEP (Python
enhancement proposals) 8 and 257. This is also inspected using lint checkers
such as \codeinline{flake8} and \codeinline{pylint} among the developers.  The
code is regularly cleaned up using the Python code formatter \codeinline{black}.

\section*{(2) Availability}
\vspace{0.5cm}
\section*{Operating system}

% \textcolor{blue}{Please include minimum version compatibility.}

Windows and any POSIX based OS, such as GNU/Linux and macOS.

\section*{Programming language}

% \textcolor{blue}{Please include minimum version compatibility.}

Python 2.7, 3.4 or above.

% \section*{Additional system requirements}

% \textcolor{blue}{E.g. memory, disk space, processor, input devices, output devices.}

\section*{Dependencies}

% \textcolor{blue}{E.g. libraries, frameworks, incl. minimum version
% compatibility.}

We list here only the dependencies of the base package \fluidpack{dyn}.

\begin{itemize}
\item {\bf Minimum:} \Numpy, \pack{Matplotlib}, \pack{psutil}, \pack{future},
\pack{subprocess32} (for Python 2.7 only).

\item {\bf Full functionality:} \pack{h5py}, \pack{h5netcdf}, \pack{mpi4py},
\pack{Scipy}, \pack{pyfftw} (requires FFTW library), \pack{pillow}.

\item {\bf Optional:} OpenCV with Python bindings, \pack{scikit-image}.
\end{itemize}
\section*{List of contributors}

% \textcolor{blue}{Please list anyone who helped to create the software (who may
% also not be an author of this paper), including their roles and affiliations.}

\begin{itemize}
\item Pierre Augier (LEGI): creator of the FluidDyn project, developer of
majority of the FluidDyn packages, future-proofing with Python 3 compatibility
and documentation.

\item Ashwin Vishnu Mohanan (KTH): developer of the packages
\fluidpack{sim}, \fluidpack{fft}, \fluidpack{image} and
\fluiddyn; documentation, code coverage and continuous integration (Docker,
Bitbucket pipeline and Travis).

\item Cyrille Bonamy (LEGI): developer of \fluidpack{fft}, \fluidpack{sim} and
\fluidpack{foam}. Main maintainer of \fluidpack{foam}.

\item Antoine Campagne (LEGI): developer of \fluidpack{image} and
\fluidpack{lab}.

\item Miguel Calpe Linares (LEGI): developer of \fluidpack{sim} and
\fluidpack{lab}.

\item Julien Salort (Laboratoire de physique, ENS de Lyon): developer of
\fluidpack{lab}.

\end{itemize}
%
% \section*{Software location:}
%
% % {\bf Archive} \textcolor{blue}{(e.g. institutional repository, general
% % repository) (required – please see instructions on journal website for
% % depositing archive copy of software in a suitable repository)}
%
% \begin{description}[noitemsep,topsep=0pt]
% \item[Name:] PyPI
% \item[Persistent identifier:] https://pypi.org/project/fluiddyn
% \item[Licence:] CeCILL-B, a BSD compatible French licence.
% \item[Publisher:] Pierre Augier
% \item[Version published:] 0.2.3
% \item[Date published:] 02/05/2018
% \end{description}
%
% {\bf Code repository}
% % \textcolor{blue}{(e.g. SourceForge, GitHub etc.) (required)}
%
% \begin{description}[noitemsep,topsep=0pt]
% \item[Name:] Bitbucket
% \item[Persistent identifier:] https://bitbucket.org/fluiddyn/fluiddyn
% \item[Licence:] CeCILL-B
% \item[Date published:] 2015
% \end{description}

% {\bf Emulation environment} \textcolor{blue}{(if appropriate)}

% \begin{description}[noitemsep,topsep=0pt]
% \item[Name:] \textcolor{blue}{The name of the archive.}
% \item[Persistent identifier:] \textcolor{blue}{e.g. DOI, handle, PURL, etc.}
% \item[Licence:] \textcolor{blue}{Open license under which the software is licensed.}
% \item[Date published:] \textcolor{blue}{dd/mm/yy}
% \end{description}

% \section*{Language}

% \textcolor{blue}{Language of repository, software and supporting files.}

% English

\section*{(3) Reuse potential}
%
% \textcolor{blue}{Please describe in as much detail as possible the ways in
% which the software could be reused by other researchers both within and outside
% of your field. This should include the use cases for the software, and also
% details of how the software might be modified or extended (including how
% contributors should contact you) if appropriate. Also you must include details
% of what support mechanisms are in place for this software (even if there is no
% support).}
%

As a library, \fluiddyn\ has been used in the project's specialized packages.
The common code base for packages with such varied applications is a proof of
\fluiddyn's versatility and generality. It can be used in other packages outside
the FluidDyn project easily depending on the need. The command-lines tools can
be useful to all scientists working with Python. Other use cases could be:
\begin{itemize}
	\item \codeinline{ParamContainer} as a generic parameter storage
		standard.
	\item Modules within \codeinline{fluiddyn.util} for miniature tasks,
		printing in terminal with colours, getting memory usage
		information, detecting if the current Python session is inside
		IPython, creating a string with time and date, etc.
	\item Modules within \codeinline{fluiddyn.io} to read
		and save images of various formats including TIFF files with
		multiple images; easily ask user with yes/no queries; handle
		different file formats using classes, including CSV, HDF5,
		Digiflow, Dantec formats.
	\item Modules inside \codeinline{fluiddyn.clusters} subpackage to
		make job submission scriptable in HPC clusters with OAR or
		SLURM job schedulers. It could be even used for non-Python
		jobs.
\end{itemize}

There is no formal support mechanism. However, bug reports can be submitted at the
\href{https://bitbucket.org/fluiddyn/fluiddyn/issues}{Issues page on Bitbucket}.
Discussions and questions can be aired on instant messaging channels in Riot (or
equivalent with Matrix protocol) at
\url{%
  https://matrix.to/#/#fluiddyn-users:matrix.org}
or via IRC protocol on Freenode at \codeinline{\#fluiddyn-users}.

\subsection*{Conclusions}

FluidDyn is an attempt to set off collaborative dynamics based on open-source
development in fluid dynamics research.
%
We shall try, with this project, to explore the possibilities of open-source in
science and fluid dynamics, by fully exploiting the new open-source tools and
methods presented in the first subsections of the introduction.

% todo: again mention some success of community-driven subject-area research
% libraries? See the file ./communitydriven_research_soft.md

The project is right now in a very preliminary stage.  Packages are actively
evolving with interesting features and a framework for collaborative
development, packaging, documenting and testing is now well set.
%
However, the community around the project is currently tiny and now, we have to
work on attracting users and developers since an active community is a
criteria for success and sustenance of an open-source project.

Will people use the FluidDyn tools and collaborate through the project
FluidDyn?
%
There are clearly many challenges and potential barriers:
\begin{itemize}
\item Some habits in the community.
\item Knowledge and skills in the community. For example, only few people use
issue tracker and pull requests.
% from Julien Salort: bad!
% \item Misunderstanding between colleagues and in particular colleagues with
% responsibilities and power.
\item Lack of a business model for open-source software in science.  A good
quality open-source software has a cost that institutions should be willing
to fund and support.
\item Lack of recognition of the work spent in open-source, in particular for
precarious scientists.
\end{itemize}

On our side, we also have positive points.
%
The quality of the tools we use (Python and its scientific ecosystem,
Mercurial, Read the Docs, Jupyter, ...) is impressive.
%
Scientific code is done to be read and to transmit ideas. To this effect, Python
is among the best languages today.
%
We can benefit from the dynamics of Python and of emerging subjects like deep
learning and the Internet of Things.
%
Finally, if we manage to gather a community of users and of developers, the
collective efficiency related to open-source methods and tools can be a strong
booster.

\section*{Acknowledgements}

% \textcolor{blue}{Please add any relevant acknowledgements to anyone else who
% supported the project in which the software was created, but did not work
% directly on the software itself.}

We thank the CNRS to finance the work of Pierre Augier while giving freedom in
terms of scientific project.
%
Similarly, Ashwin Vishnu Mohanan could not have been as involved in this
project without the kindness of Erik Lindborg.
%
We thank Cyrille Bonamy, Antoine Campagne, Miguel Calpe Linares and Julien
Salort for their implications in the development of some of the FluidDyn
packages.
%
We thank Bitbucket for providing us for free a high quality forge compatible with
Mercurial.
%
Finally, we thank Gabriel Moreau and Jo\"el Sommeria for their constant willing
to discuss and share their knowledge.

\section*{Funding statement}

% \textcolor{blue}{If the software resulted from funded research please give the
% funder and grant number.}

This project has indirectly benefit from funding from the foundation Simone et
Cino Del Duca de l'Institut de France, the European Research Council (ERC)
under the European Union's Horizon 2020 research and innovation program (grant
agreement No 647018-WATU and Euhit consortium) and the Swedish Research Council
(Vetenskapsr{\aa}det): 2013-5191.
%
We have also been able to use supercomputers of CIMENT/GRICAD, CINES/GENCI and
the Swedish National Infrastructure for Computing (SNIC).

% \section*{Competing interests}

% \textcolor{blue}{If any of the authors have any competing interests then these
% must be declared. The authors’ initials should be used to denote differing
% competing interests. For example: “BH has minority shares in [company name],
% which part funded the research grant for this project. All other authors have
% no competing interests."
% %
% If there are no competing interests, please add the statement: “The authors
% declare that they have no competing interests.” }

% The authors declare that they have no competing interests.

% already included by \bibliography{../bib}
% \section*{References}

% \textcolor{blue}{Please enter references in the Harvard style and include a DOI
% where available, citing them in the text with a number in square brackets,
% e.g.}

% \textcolor{blue}{[1] Piwowar, H A 2011 Who Shares? Who Doesn't? Factors
% Associated with Openly Archiving Raw Research Data. PLoS ONE 6(7): e18657. DOI:
% \\ http://dx.doi.org/10.1371/journal.pone.0018657.}
%
% \bibliographystyle{../agsm}
% \bibliography{../bib}

% \rule{\textwidth}{1pt}
%
% { \bf Copyright Notice} \\
% Authors who publish with this journal agree to the following terms: \\
%
% Authors retain copyright and grant the journal right of first publication with
% the work simultaneously licensed under a
% \href{http://creativecommons.org/licenses/by/3.0/}{Creative Commons Attribution
% License} that allows others to share the work with an acknowledgement of the
% work's authorship and initial publication in this journal.
%
% Authors are able to enter into separate, additional contractual arrangements
% for the non-exclusive distribution of the journal's published version of the
% work (e.g., post it to an institutional repository or publish it in a book),
% with an acknowledgement of its initial publication in this journal.
%
% By submitting this paper you agree to the terms of this Copyright Notice, which
% will apply to this submission if and when it is published by this journal.
%
%

>>>>

\chapter{Towards reproducible, open science through open research software}
%TODO: Fix table of contents
\section{Introduction}

% \vspace{0.5cm}

% \section*{Title}

@@

% Taking advantage of recent advances in open-source software engineering to study
% fluid dynamics: the example of the FluidDyn project

@@

\subsection{Science, software, open-source and the computer revolution}

@@

%TODO: Fix names
The section ``Introduction'' is completed by a second subsection called
``Implementation and architecture'' devoted to present the organization of the
packages of the project and to describe in more detail the base package
\fluiddyn\footnote{We use FluidDyn (with capital letters) to name the project and
\fluiddyn for the base package.}.

\subsection{Methods and tools for open-source software engineering}

@@

\subsection{Python, a programming language adapted for open-science}

@@

\paragraphbf{An incredible success}

@@

\paragraphbf{Some Python issues}

@@

\paragraphbf{Rich landscape of open-source languages useful for science}

@@

\paragraphbf{Summary: Python, a versatile Swiss Army Knife for the
scientists}

@@

\subsection{Productivity at individual, group and community levels}

@@

\subsection{Programming in the field of fluid mechanics}

@@

\paragraphbf{Discussion on Matlab, Labview and Mathematica}

@@

\paragraphbf{Different models for software development in fluid mechanics}

@@

\section{FluidDyn project: Breaking the status quo}
% \section{Implementation and architecture}

@@

@@

\subsection{Organization of the code in packages}

@@

\subsection{API of the Python library \fluiddyn}

@@

Sub-packages \codeinline{io}, \codeinline{util} and \codeinline{calcul} are the
largest in terms of lines of code and provides the Application Programming
Interfaces (API) to support \fluidpack{sim}, \fluidpack{lab} and
\fluidpack{image}.

\section{Conclusion}

% FluidDyn is an attempt to set off collaborative dynamics based on open-source
% development in fluid dynamics research.
% %
% We shall try, with this project, to explore the possibilities of open-source in
% science and fluid dynamics, by fully exploiting the new open-source tools and
% methods presented in the first subsections of the introduction.
%
% % todo: again mention some success of community-driven subject-area research
% % libraries? See the file ./communitydriven_research_soft.md
%
% The project is right now in a very preliminary stage.  Packages are actively
% evolving with interesting features and a framework for collaborative
% development, packaging, documenting and testing is now well set.
% %
% However, the community around the project is currently tiny and now, we have to
% work on attracting users and developers since an active community is a
% criteria for success and sustenance of an open-source project.
%
% Will people use the FluidDyn tools and collaborate through the project
% FluidDyn?
% %
% There are clearly many challenges and potential barriers:
% \begin{itemize}
% \item Some habits in the community.
% \item Knowledge and skills in the community. For example, only few people use
% issue tracker and pull requests.
% % from Julien Salort: bad!
% % \item Misunderstanding between colleagues and in particular colleagues with
% % responsibilities and power.
% \item Lack of a business model for open-source software in science.  A good
% quality open-source software has a cost that institutions should be willing
% to fund and support.
% \item Lack of recognition of the work spent in open-source, in particular for
% precarious scientists.
% \end{itemize}
%
% On our side, we also have positive points.
% %
% The quality of the tools we use (Python and its scientific ecosystem,
% Mercurial, Read the Docs, Jupyter, ...) is impressive.
% %
% Scientific code is done to be read and to transmit ideas. To this effect, Python
% is among the best languages today.
% %
% We can benefit from the dynamics of Python and of emerging subjects like deep
% learning and the Internet of Things.
% %
% Finally, if we manage to gather a community of users and of developers, the
% collective efficiency related to open-source methods and tools can be a strong
% booster.
